BOLD Parameter Estimation Using Particle Filters

### 2010-09-02 doc/thesis/background.tex
fixed all the acronyms, think I might be done

\begin{figure}
\centering
\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_point1sec_3sigma.pdf}
\caption[Q-Q Plot Of Standard Normal and Result of $0.1$ s BOLD Simulation Output]
{Q-Q Plot Of Standard Normal and Result of $0.1$ s BOLD Simulation Output. A
perfect Gaussian would fall on the green line.}
\label{fig:transp1s}
\end{figure}

\begin{figure}
\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_1sec_3sigma.pdf}
\caption[Q-Q Plot Of Standard Normal and Result of $1$ s BOLD Simulation Output]
{Q-Q Plot Of Standard Normal and Result of $1$ s BOLD Simulation Output. A
perfect Gaussian would fall on the green line.}
\label{fig:trans1s}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_point1sec_3sigma.pdf}
\caption[Q-Q Plot Of Standard Normal and Result of $0.1$ s \ac{BOLD} Simulation Output]
{Q-Q Plot Of Standard Normal and Result of $0.1$ s \ac{BOLD} Simulation Output. A
perfect Gaussian would fall on the green line.}
\label{fig:transp1s}
\end{figure}

\begin{figure}
\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_1sec_3sigma.pdf}
\caption[Q-Q Plot Of Standard Normal and Result of $1$ s \ac{BOLD} Simulation Output]
{Q-Q Plot Of Standard Normal and Result of $1$ s \ac{BOLD} Simulation Output. A
perfect Gaussian would fall on the green line.}
\label{fig:trans1s}
\end{figure}

### 2010-09-02 doc/thesis/intro.tex
fixed all the acronyms, think I might be done

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{images/bold}
\caption[Diagram of BOLD Signal]{A) End of Pre-Stimulus Undershoot, $< 2s$ typical,
B) End of Stimulus, C) Input Bloodflow Drops from Supernormal to Subnormal,
$5 s$ after B typical, D) Beginning of Post-stimulus Undershoot, $12 s$ after B
typical, E) Full return to normal, between $20-30s$ after B. \cite{Behzadi2005}}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{images/bold}
\caption[Diagram of \ac{BOLD} Signal]{A) End of Pre-Stimulus Undershoot, $< 2s$ typical,
B) End of Stimulus, C) Input Bloodflow Drops from Supernormal to Subnormal,
$5 s$ after B typical, D) Beginning of Post-stimulus Undershoot, $12 s$ after B
typical, E) Full return to normal, between $20-30s$ after B. \cite{Behzadi2005}}
\end{figure}

### 2010-09-02 doc/thesis/methods.tex
fixed all the acronyms, think I might be done

Problematically, after removing the DC component of the signal,
by definition the signal will have a median near zero.
Unfortunately this is not the natural state of the \ac{BOLD} signal. More specifically,
when the signal is inactive, the \ac{BOLD} response should be at 0\% change from
the base level; activation may then increase, or for short periods decrease from this base.
Because most of the \ac{BOLD} signal is above baseline, after removing the spline
the resting state will be below 0\%.
One method of accounting for this is to simply add a DC gain model parameter.
Like all the other model parameters, with enough measurements, the particle filter
would be able to settle on a good estimate. Yet adding another dimension increases the
complexity of the model, for a parameter that is relatively easy to estimate
by visual inspection.  In this work a simpler approach was used. To determine
the DC gain a robust estimator of scale was used. The 
\ac{MAD} proved to be accurate in determining how much to shift the signal up
by. Both methods analysis were tested, and it was found that the increase
in model complexity far outweighed the slight increase in flexibility. Other
methods may work better, however the \ac{MAD} worked well,
as \autoref{fig:PreprocessedLowNoise} and 
\autoref{fig:PreprocessedHighNoise} show. The \ac{DC} gain was then set to:

@@

For more information on the \ac{MAD}, it is discussed in great detail in
Iglewicz's text on robust estimation \cite{iglewicz1983robust}.
A serious concern when adding constant values to
real data is whether this will create false positives. This is a legitimate
concern; however, a boosted response does not affect how well the \ac{BOLD} model
predicts the actual measurements; and as mentioned before, the DC signal 
of \ac{fMRI} is never used.

>>>>

Problematically, after removing the \ac{DC} component of the signal,
by definition the signal will have a median near zero.
Unfortunately this is not the natural state of the \ac{BOLD} signal. More specifically,
when the signal is inactive, the \ac{BOLD} response should be at 0\% change from
the base level; activation may then increase, or for short periods decrease from this base.
Because most of the \ac{BOLD} signal is above baseline, after removing the spline
the resting state will be below 0\%.
One method of accounting for this is to simply add a \ac{DC} gain model parameter.
Like all the other model parameters, with enough measurements, the particle filter
would be able to settle on a good estimate. Yet adding another dimension increases the
complexity of the model, for a parameter that is relatively easy to estimate
by visual inspection.  In this work a simpler approach was used. To determine
the \ac{DC} gain a robust estimator of scale was used. The 
\ac{MAD} proved to be accurate in determining how much to shift the signal up
by. Both methods analysis were tested, and it was found that the increase
in model complexity far outweighed the slight increase in flexibility. Other
methods may work better, however the \ac{MAD} worked well,
as \autoref{fig:PreprocessedLowNoise} and 
\autoref{fig:PreprocessedHighNoise} show. The \ac{DC} gain was then set to:

@@

For more information on the \ac{MAD}, it is discussed in great detail in
Iglewicz's text on robust estimation \cite{iglewicz1983robust}.
A serious concern when adding constant values to
real data is whether this will create false positives. This is a legitimate
concern; however, a boosted response does not affect how well the \ac{BOLD} model
predicts the actual measurements; and as mentioned before, the \ac{DC} signal 
of \ac{fMRI} is never used.

### 2010-09-02 doc/thesis/particlefilters.tex
just need to fix figure captions for multivoxel cases. And fix bib
fixed all the acronyms, think I might be done

\chapter{Particle Filters}
\label{sec:Particle Filter}
\section{Introduction}
Particle filters, a type of Sequential Monte Carlo (SMC) method,
are a powerful method for estimating the posterior probability distribution
 of parameters given a timeseries of measurements. Unlike Markov 
Chain Monte Carlo (MCMC) estimation, particle filters are designed for 
time-varying random variables. The idea behind particle filters is
similar to Kalman Filters; however, unlike Kalman Filters,
distributions are stored as an empircal distribution rather than 
as the first two moments of a Gaussian. Thus, particle filters are 
preferred when the model is nonlinear, and non-gaussian. This section
is largely based on Arulampalam et al. and Thrun et al. \cite{Arulampalam2002a, Thrun2005}.

@@

I will also make a few  simplifying assumptions for this work. 
First, the systems are assumed to be 
time invariant. This 
assumption is based on the idea that if you paused the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happened. 
Few biological systems are predictable enough for them to be summarized
by a time varying function, least of all the brain. While heart beats are certainly
periodic and have an effect on the BOLD signal, the period varies too much
for the system to be considered varying with time. 
Next, it was assumed that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
Finally, because the only difference between the members of $X(t)$ and 
$\Theta$ is an update function, from now on $X$ will encapsulate
$\Theta$. The assumptions allow for a simplified version of the
state space equations:

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$, however in this work it is a good assumption because the
input does not immediately effect the BOLD signal. 
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
This choice for importance density is conventient since its simply 
the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen later in \autoref{eq:weightevolve}.

@@

As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), a significant proportion of particles will quickly
develop insignificant weights. 
While this does help
describe the tails of the distribution, it means a great deal of computation will be wasted.
Instead, it would be preferable if most of the computation is spent on the most probable regions.
Ideally the computation time spent on tails would be proportional to the actual size of the
tails. In this case particle locations would match the true posterior and all weights would
be equal.  The case where a large number of the weights have become extremely small
is called particle degeneracy. In  Lui et al. 
an ideal calculation of the effective number of particles is found based on the 
particles' true weight \cite{Liu1998b}. However, given that only an approximation 
for the true weight 
exists, they also provide a simple heuristic calculation of $N_{eff}$.
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
Unless the prior is particularly accurate,
$N_{eff}$ drops very quickly.  To alleviate particle degeneracy
a common technique known as resampling is often applied. The idea of resampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a better support. Therefore, a new set of particles may be drawn from the empirical
distribution as follows:
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}

@@

Where $h$ is an optional bandwidth, $\sigma$ is the standard deviation such that 
$\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel. The choice of a kernel is
complex, although it has been proven that the optimal kernel for reducing MSE between
the original and resampled distributions is 
is the Epanechnikov Kernel \cite{Musso2001a}. However, Musso et al. 
also espoused the usefulness of the Gaussian Kernel, due to the ease
drawing samples from it, which for this work was more important \cite{Musso2001a}.

@@

Hurzeler et al. demonstrated that if the underlying 
distribution is non-Gaussian, then using the original bandwidth may over-smooth
\cite{Hurzeler1998}.  In reality, over smoothing
will only become an issue if resampling is performed often. Thus
if resampling is performed at every step then this would certainly cause problems.
If the distribution is over-smoothed then the algorithm may not converge as rapidly,
or at all. However, because the bandwidth is still based on particle variance, 
which should decay as particles are ruled out, the particle filter is still able to converge. 
In fact, over-smoothing is preferable
to under smoothing, since over-smoothing simply slows convergence while 
under-smoothing could leave gaps in the distribution.
Moreover, because of the high dimensionality of the BOLD model,
and limited measurements, it is helpful to have a broader bandwidth to explore the distribution.

@@

%A half wave rectifier takes a AC voltage circuit and removes

>>>>

\chapter{Particle Filters}
\label{sec:Particle Filter}
\section{Introduction}
Particle filters, a type of Sequential Monte Carlo (SMC) method,
are a powerful method for estimating the posterior probability distribution
 of parameters given a timeseries of measurements. Unlike Markov 
Chain Monte Carlo (MCMC) estimation, particle filters are designed for 
time-varying random variables. The idea behind particle filters is
similar to Kalman Filters; however, unlike Kalman Filters,
distributions are stored as an empircal distribution rather than 
as the first two moments of a Gaussian. Thus, particle filters are 
preferred when the model is nonlinear, and non-gaussian. This section
is largely based on Arulampalam et al. and Thrun et al. \cite{Arulampalam2002a, 
thrun2008probabilistic}.

@@

I will also make a few  simplifying assumptions for this work. 
First, the systems are assumed to be 
time invariant. This 
assumption is based on the idea that if you paused the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happened. 
Few biological systems are predictable enough for them to be summarized
by a time varying function, least of all the brain. While heart beats are certainly
periodic and have an effect on the \ac{BOLD} signal, the period varies too much
for the system to be considered varying with time. 
Next, it was assumed that input cannot directly
influence the output, which in the case of the \ac{BOLD} signal is a good assumption.
Finally, because the only difference between the members of $X(t)$ and 
$\Theta$ is an update function, from now on $X$ will encapsulate
$\Theta$. The assumptions allow for a simplified version of the
state space equations:

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$, however in this work it is a good assumption because the
input does not immediately effect the \ac{BOLD} signal. 
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
This choice for importance density is conventient since its simply 
the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen later in \autoref{eq:weightevolve}.

@@

As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), a significant proportion of particles will quickly
develop insignificant weights. 
While this does help
describe the tails of the distribution, it means a great deal of computation will be wasted.
Instead, it would be preferable if most of the computation is spent on the most probable regions.
Ideally the computation time spent on tails would be proportional to the actual size of the
tails. In this case particle locations would match the true posterior and all weights would
be equal.  The case where a large number of the weights have become extremely small
is called particle degeneracy. In  Lui et al. 
an ideal calculation of the effective number of particles is found based on the 
particles' true weight \cite{Liu1998}. However, given that only an approximation 
for the true weight 
exists, they also provide a simple heuristic calculation of $N_{eff}$.
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
Unless the prior is particularly accurate,
$N_{eff}$ drops very quickly.  To alleviate particle degeneracy
a common technique known as resampling is often applied. The idea of resampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a better support. Therefore, a new set of particles may be drawn from the empirical
distribution as follows:
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}

@@

Where $h$ is an optional bandwidth, $\sigma$ is the standard deviation such that 
$\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel. The choice of a kernel is
complex, although it has been proven that the optimal kernel for reducing \ac{MSE} between
the original and resampled distributions is 
is the Epanechnikov Kernel \cite{Musso2001a}. However, Musso et al. 
also espoused the usefulness of the Gaussian Kernel, due to the ease
drawing samples from it, which for this work was more important \cite{Musso2001a}.

@@

Hurzeler et al. demonstrated that if the underlying 
distribution is non-Gaussian, then using the original bandwidth may over-smooth
\cite{Hurzeler1998}.  In reality, over smoothing
will only become an issue if resampling is performed often. Thus
if resampling is performed at every step then this would certainly cause problems.
If the distribution is over-smoothed then the algorithm may not converge as rapidly,
or at all. However, because the bandwidth is still based on particle variance, 
which should decay as particles are ruled out, the particle filter is still able to converge. 
In fact, over-smoothing is preferable
to under smoothing, since over-smoothing simply slows convergence while 
under-smoothing could leave gaps in the distribution.
Moreover, because of the high dimensionality of the \ac{BOLD} model,
and limited measurements, it is helpful to have a broader bandwidth to explore the distribution.

@@

%A half wave rectifier takes a \ac{AC} voltage circuit and removes
